\bf Properties and Bayesian fitting of restricted Boltzmann machines

### 2018-09-04 writing/paper-rbm/docs/paper.tex
more wording changes

First, the models can be unsatisfactory as conceptualizations of how
data are generated. That is, recalling Fisher
(\protect\hyperlink{ref-fisher1922mathematical}{1922}), the aim of a
statistical model is to represent data in a compact way. Neyman and Box
further state that a model should ``provide an explanation of the
mechanism underlying the observed phenomena'' (Lehmann
\protect\hyperlink{ref-lehmann1990model}{1990}; G. E. P. Box
\protect\hyperlink{ref-box1967discrimination}{1967}). At issue,
simulation from RBMs can often produce data lacking realistic
variability so that such models may thereby fail to satisfactorily
reflect a true data generation process. Such behavior relates to model
degeneracy (or near degeneracy), which is a statistical concern in that
most data processes being modeled are realistically not degenerate in
their spectrum of potential outcomes. For example, when sampling data
from a nearly degenerate RBM, only a small number of output
possibilities receives nearly all probability, and thus a sample of
images will all be copies of the same one image (or small number of
images). An example of ten 4-pixel images simulated from a nearly
degenerate RBM model is compared to ten 4-pixel images simulated from a
non-degenerate RBM model in Figure \ref{fig:sample-models}. The
degenerate model places almost all probability on one outcome, causing
the image to be generated repeatedly, whereas the non-degenerate model
shows more potentially realistic variation.

\par
\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-1} \includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-2} \caption{Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b). The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.}\label{fig:sample-models}
\end{figure}
In addition to such degeneracy, we find that RBMs can easily exhibit
types of instability, related to how sensitive the model probability
structure can be to small changes in data outcomes. In practice, this
may be seen when a single pixel change in an image results in a wildly
different classification in an image classification problem. Occurrences
of such behavior have recently been documented in RBMs (Li
\protect\hyperlink{ref-li2014biclustering}{2014}), as well as other deep
architectures (Szegedy et al.
\protect\hyperlink{ref-szegedy2013intriguing}{2013}; Nguyen, Yosinski,
and Clune \protect\hyperlink{ref-nguyen2014deep}{2014}). We describe
potential issues of model instability, degeneracy and uninterpretability
for the RBM class (which are related properties) and examine the
presence of these in Section
\ref{explorations-of-model-properties-through-simulation} through
simulations of small, manageable examples.

Separate of the exploration of model properties, we also investigate the
quality of estimation in fitting these models and examine an estimation
procedure that allows for uncertainty quantification rather than point
estimates alone. The fitting of RBMs can be problematic for two reasons,
the first being computational and the second concerning flexibility. As
the size of these models grows, both maximum likelihood and Bayesian
methods of fitting quickly become intractable. The literature often
suggests Markov chain Monte Carlo (MCMC) tools for approximate
maximization of likelihoods to fit these models (e.g., Gibbs sampling to
exploit conditional structure in hidden and visible variables), but
little is said about the attributes of realized estimates (Hinton
\protect\hyperlink{ref-hinton2010practical}{2010}; Hinton, Osindero, and
Teh \protect\hyperlink{ref-hinton2006fast}{2006}), including the absence
of interval estimates.

Furthermore, these MCMC algorithms require updating potentially many
latent variables (hiddens) which can critically influence convergence in
MCMC-based likelihood methods. Applying basic statistical principles in
fitting RBM models of tractable sizes, we compare three fully Bayesian
techniques involving MCMC, which are computationally more accessible
than direct maximum likelihood and also aim to avoid parts of a RBM
parameter space that yield unattractive models. As might be expected,
with greater computational complexity comes an increase in fitting
accuracy, but at the cost of practical feasibility.

As a compounding factor to the computational challenges, issues in model
fitting due to model flexibility are potentially more concerning. For a
RBM model with enough hidden variables, it has been shown that any
distribution for the visibles can be approximated arbitrarily well (Le
Roux and Bengio \protect\hyperlink{ref-le2008representational}{2008};
Montufar and Ay \protect\hyperlink{ref-montufar2011refinements}{2011};
and Montúfar, Rauh, and Ay
\protect\hyperlink{ref-montufar2011expressive}{2011}). However, for
binary data, the empirical distribution of an observed training set of
visible variables provides a highest likelihood benchmark, before any
parametric model class is even introduced and applied to obtain a
refinement of model fit. As a consequence, we find that any fully
principled fitting method based on the likelihood for a RBM with enough
hidden variables will seek to reproduce the (discrete) empirical
distribution of a training set. This aspect can be undesirable, and
perhaps even unexpected compared to most modeling scenarios, in that no
``smoothed distribution'' may result when fitting a RBM model of
sufficient size with a rigorous likelihood-based method. We are
therefore led to be skeptical that models that involve these structures
(like deep Boltzmann machines) can achieve useful prediction or
inference in a principled way without intentionally limiting the
flexibility of the fitted model.

Notions of weight-decay (penalization) and sparsity (regularization)
have been suggested in the RBM literature as practical remedies for both
over-fitting and poor mixing in the Markov chain during fitting (Hinton
\protect\hyperlink{ref-hinton2010practical}{2010}; Tieleman
\protect\hyperlink{ref-tieleman2008training}{2008}; Cho, Ilin, and Raiko
\protect\hyperlink{ref-cho2012tikhonov}{2012}). Both \(L_1\) and \(L_2\)
type penalties are mentioned to achieve weight-decay, though the
benefits of these particular forms are unknown in different situations.
However, the degree to which regularization and penalization are used by
practitioners is not clear because these concepts are not (by default) a
part of the overall model-specification-and-fitting methodology (Hinton
\protect\hyperlink{ref-hinton2002training}{2002}; Carreira-Perpinan and
Hinton \protect\hyperlink{ref-carreira2005contrastive}{2005}). In this
paper, we attempt to address the concerns with overfitting and poor
mixing in an alternative (and perhaps more transparent) manner, via
specification of a Bayesian prior in Section
\ref{bayesian-model-fitting}.

This paper is structured as follows. Section \ref{background} formally
defines the RBM including the joint distribution of hidden and visible
variables and explains the model's connection to deep learning.
Additionally, measures of model impropriety and methods of
quantifying/detecting it are defined. Section
\ref{explorations-of-model-properties-through-simulation} details our
explorations into the model behavior and potential propriety issues with
the RBM class. This aligns with the call to action in Sculley et al.
(\protect\hyperlink{ref-sculley2018winner}{2018}) to employ small toy
examples to advance empirical understanding of deep learning models and
strive for a ``higher level of empirical rigor in the field.'' We
examine three Bayesian fitting techniques intended to avoid model
impropriety issues raised in Section \ref{model-fitting} and conclude
with a discussion in Section \ref{discussion}. Supplementary online
material provides proofs for results on RBM parameterizations and data
codings described in Section \ref{data-encoding}.

While applications of the RBM have been claimed to produce some
successes in classification problems, it is unclear if the model class
allows one to go beyond fitting to other statistical matters of
importance in using the models, such as quantification of the
uncertainty of estimation and the formulation of predictive
distributions. These are closely tied to the probability properties of
the RBM model for explaining data generation and the current exposition
intends to contribute to a better understanding in this direction.

@@

Let \(\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)\) represent
the states of the visible and hidden nodes in a RBM for some integers
\(\nv, \nh \ge 1\). Each single binary random variable, visible or
hidden, will take its values in a common coding set \(\mathcal{C}\),
where we allow one of two possibilities for the coding set,
\(\mathcal{C}=\{0,1\}\) or \(\mathcal{C}=\{-1,1\}\), with ``\(1\)''
always indicating the ``high'' value of the variable. While
\(\mathcal{C}=\{0,1\}\) may be a natural starting point, we argue in
Section \ref{explorations-of-model-properties-through-simulation} that
the coding \(\mathcal{C}=\{-1,1\}\) induces more interpretable model
properties for the RBM. A standard parametric form for probabilities
corresponding to a potential vector of states,
\(X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)\), for the nodes is
\begin{align}
\label{eq:pmf}
f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv} 
\end{align} where
\(\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}\)
denotes the vector of model parameters and the denominator \[
\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
\] is the normalizing function that ensures the probabilities
\eqref{eq:pmf} sum to one. For
\(\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}\)
and \begin{align}
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eq:t}
\end{align} let
\(\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}\)
be the set of possible values for the vector of variables needed to
compute probabilities \eqref{eq:pmf} in the model, and write
\(Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j\)
for the ``neg-potential'' function. The RBM model is parameterized by
\(\boldsymbol \theta\) containing two types of parameters, main effects
and interaction effects. The main effects parameters
(\(\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}\))
weight the values of the visible \(v_i\) and hidden \(h_j\) nodes in
probabilities \eqref{eq:pmf} and the interaction effect parameters
(\(\theta_{ij}\)) weight the values of the connections \(v_i h_j\), or
dependencies, between hidden and visible layers.

@@

RBMs have risen to prominence in recent years due to their connection to
deep learning (see Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}; Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2012efficient}{2012}; Srivastava,
Salakhutdinov, and Hinton
\protect\hyperlink{ref-srivastava2013modeling}{2013} for examples). By
stacking multiple layers of RBMs in a deep architecture, proponents of
the model are claimed to produce the ability to learn ``internal
representations that become increasingly complex, which is considered to
be a promising way of solving object and speech recognition problems''
(Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009}, 450). The stacking
is achieved by treating a hidden layer of one RBM as the visible layer
in a second RBM, and so on, until the desired multi-layer architecture
is created.

@@

In Random Graph Model theory, \emph{model degeneracy} means there is a
disproportionate amount of probability placed on only a few elements of
the sample space, \(\mathcal{X}\), by the model (Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). For random graph
models, \(\mathcal{X}\) denotes all possible graphs that can be
constructed from a set of nodes and an exponentially parameterized
random graph model has a distribution with a probability mass function
of the form \begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*} where
\(\boldsymbol \theta \in \Theta \subset \mathbb{R}^q\) is the model
parameter, and \(\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q\)
is a vector of statistics based on the adjacency matrix of a graph.
Here, as earlier,
\(\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)\)
is the normalizing function. Let \(C\) denote the convex hull of the
potential outcomes of sufficient statistics,
\(\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}\),
under the model above. Handcock
(\protect\hyperlink{ref-handcock2003assessing}{2003}) classifies an
exponentially parameterized random graph model at \(\boldsymbol \theta\)
as \emph{near-degenerate} if the mean value of the vector of sufficient
statistics under \(\boldsymbol \theta\),
\(\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\),
is close to the boundary of \(C\). Intuitively, if a model is
near-degenerate in the sense that only a small number of elements of the
sample space \(\mathcal{X}\) have positive probability, the expected
value \(\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\) is
an average of that same small number of values of \(t( \boldsymbol x)\)
(defining the boundary of the hull \(C\)) and can be expected to
\emph{not} be pulled deep into the interior of \(C\).

A RBM model can be thought to exhibit an analogous form of
\emph{near-degeneracy} when there is a disproportionate amount of
probability placed on a small number of elements in the sample space of
visibles and hiddens, \(\mathcal{C}^{\nv + \nh}\). Using the idea of
Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003}), when the
random vector
\(\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}\)
from \eqref{eq:t}, appearing in the neg-potential function
\(Q_{\boldsymbol \theta}(\cdot)\), has a mean vector
\(\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}\)
close to the boundary of the convex hull of \(\mathcal{T}\), then the
RBM model can be said to exhibit near-degeneracy at
\(\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}\). Here the
mean of \(\boldsymbol t(\boldsymbol x)\) is \begin{align*}
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  &= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
\end{align*}
\hypertarget{instability}{%
\subsubsection{Instability}\label{instability}}

@@

S-unstable RBM model sequences are undesirable for several reasons. One
is that, as mentioned above, small changes in data can lead to
overly-sensitive changes in probability under the data model. Consider,
for example, \[
\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(\boldsymbol v)}{P_{\boldsymbol \theta_\nv}(\boldsymbol v^*)} : \boldsymbol v \text{ }\& \text{ } \boldsymbol v^* \in \mathcal{C}^\nv \text{ differ in exactly one component}\right\},
\] denoting the biggest log-probability ratio for a one component change
in data outcomes (visibles) at a RBM parameter
\(\boldsymbol \theta_\nv\). We then have the following result.

@@

marginal mean-structures fail to hold as anticipated by a model
statement. We adapt this notion to RBM models as follows. Note that in a

@@

destroying the model interpretation of dependence as local conditional
modifications of an overall marginal mean structure). Whether or not
parameter interpretation is itself a goal in a given application of RBM
models, this concept of interpretation can provide an additional device

>>>>

First, the models can be unsatisfactory as conceptualizations of how
data are generated. That is, recalling Fisher
(\protect\hyperlink{ref-fisher1922mathematical}{1922}), the aim of a
statistical model is to represent data in a compact way. Neyman and Box
further state that a model should ``provide an explanation of the
mechanism underlying the observed phenomena'' (Lehmann
\protect\hyperlink{ref-lehmann1990model}{1990}; G. E. P. Box
\protect\hyperlink{ref-box1967discrimination}{1967}). At issue,
simulation from RBMs can often produce data lacking realistic
variability so that such models may thereby fail to satisfactorily
reflect an observed data generation process. Such behavior relates to
model degeneracy (or near degeneracy), which is a statistical concern in
that many data processes of interest are realistically not degenerate in
their spectrum of potential outcomes. For example, when sampling data
from a nearly degenerate RBM used to model an imaging process, only a
small set of output possibilities receives nearly all probability, and
thus a sample of images will all be copies of the same one image (or
small number of images). An example of ten 4-pixel images simulated from
a nearly degenerate RBM model is compared to ten 4-pixel images
simulated from a non-degenerate RBM model in Figure
\ref{fig:sample-models}. The degenerate model places almost all
probability on one outcome, causing the image to be generated
repeatedly, whereas the non-degenerate model shows more potentially
realistic variation.

\par
\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-1} \includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-2} \caption{Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b). The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.}\label{fig:sample-models}
\end{figure}
In addition to such degeneracy, we find that RBMs can easily exhibit
types of instability, related to how sensitive the model probabilities
can be to small changes in data outcomes. In practice, this may be seen
when a single pixel change in an image results in a wildly different
classification in an image classification problem. Occurrences of such
behavior have recently been documented in RBMs (Li
\protect\hyperlink{ref-li2014biclustering}{2014}), as well as other deep
architectures (Szegedy et al.
\protect\hyperlink{ref-szegedy2013intriguing}{2013}; Nguyen, Yosinski,
and Clune \protect\hyperlink{ref-nguyen2014deep}{2014}). We describe
(related) potential issues of model instability, degeneracy and
uninterpretability for the RBM class (which are related properties) and
examine the presence of these in Section
\ref{explorations-of-model-properties-through-simulation} through
simulations of small, manageable examples.

Separate from the exploration of model properties, we also investigate
the quality of estimation possible in fitting these models and examine
an estimation procedure that allows for uncertainty quantification
rather than point estimates alone. The fitting of RBMs can be
problematic for two reasons, the first being computational and the
second concerning flexibility. As the size of these models grows, both
maximum likelihood and Bayesian methods of fitting quickly become
intractable. The literature often suggests Markov chain Monte Carlo
(MCMC) tools for approximate maximization of likelihoods to fit these
models (e.g., Gibbs sampling to exploit conditional structure in hidden
and visible variables), but little is said about the attributes of
realized estimates (Hinton
\protect\hyperlink{ref-hinton2010practical}{2010}; Hinton, Osindero, and
Teh \protect\hyperlink{ref-hinton2006fast}{2006}), including the absence
of interval estimates.

Furthermore, these MCMC algorithms involve updating potentially many
latent variables (hiddens) which can critically influence convergence in
MCMC-based likelihood methods. Applying basic statistical principles in
fitting RBM models of tractable size, we compare three fully Bayesian
techniques involving MCMC, which are computationally more accessible
than direct maximum likelihood and also aim to avoid parts of a RBM
parameter space that yield unattractive models. As might be expected,
with greater computational complexity comes an increase in fitting
accuracy, but at the cost of practical feasibility.

As a factor compounding the computational challenges, issues in model
fitting traceable to model flexibility are potentially more concerning.
For a RBM model with enough hidden variables, any distribution for the
visibles can be approximated arbitrarily well (Le Roux and Bengio
\protect\hyperlink{ref-le2008representational}{2008}; Montufar and Ay
\protect\hyperlink{ref-montufar2011refinements}{2011}; and Montúfar,
Rauh, and Ay \protect\hyperlink{ref-montufar2011expressive}{2011}).
However, for binary data, the empirical distribution of an observed
training set of visible variables provides a highest likelihood
benchmark, before any parametric model class is even introduced and
applied to obtain a refinement of model fit. As a consequence, we find
that any fully principled fitting method based on the likelihood for a
RBM with enough hidden variables will seek to reproduce the (discrete)
empirical distribution of a training set. This aspect can be
undesirable, and perhaps even unexpected compared to most modeling
scenarios, in that no ``smoothed distribution'' may result when fitting
a RBM model of sufficient size with a rigorous likelihood-based method.
We are therefore led to be skeptical that models that involve these
structures (like deep Boltzmann machines) can achieve useful prediction
or inference in a principled way without intentionally limiting the
flexibility of the fitted model.

Notions of weight-decay (penalization) and sparsity (regularization)
have been suggested in the RBM literature as practical remedies for both
over-fitting and poor mixing in the Markov chain during fitting (Hinton
\protect\hyperlink{ref-hinton2010practical}{2010}; Tieleman
\protect\hyperlink{ref-tieleman2008training}{2008}; Cho, Ilin, and Raiko
\protect\hyperlink{ref-cho2012tikhonov}{2012}). Both \(L_1\) and \(L_2\)
type penalties are mentioned to achieve weight-decay, though the
benefits of these particular forms are unknown in different situations.
The degree to which regularization and penalization are used by
practitioners is not clear because these concepts are not (by default) a
part of the standard overall model-specification-and-fitting methodology
(Hinton \protect\hyperlink{ref-hinton2002training}{2002};
Carreira-Perpinan and Hinton
\protect\hyperlink{ref-carreira2005contrastive}{2005}). In this paper,
we attempt to address the concerns with overfitting and poor mixing in
an alternative (and perhaps more transparent) manner, via specification
of a Bayesian prior in Section \ref{bayesian-model-fitting}.

This paper is structured as follows. Section \ref{background} formally
defines the RBM including the joint distribution of hidden and visible
variables and explains the model's connection to deep learning.
Additionally, measures of model impropriety and methods of
quantifying/detecting it are defined. Section
\ref{explorations-of-model-properties-through-simulation} details our
explorations into the model behavior and potential propriety issues with
the RBM class. This work aligns with the call to action by Google
scientists in Sculley et al.
(\protect\hyperlink{ref-sculley2018winner}{2018}) to employ small toy
examples to advance empirical understanding of deep learning models and
strive for a ``higher level of empirical rigor in the field.'' We
examine three Bayesian fitting techniques intended to avoid model
impropriety issues raised in Section \ref{model-fitting} and conclude
with a discussion in Section \ref{discussion}. Supplementary online
material provides proofs for results on RBM parameterizations and data
codings described in Section \ref{data-encoding}.

While applications of the RBM have been claimed to produce some
successes in classification problems, it is unclear if the model class
allows one to go beyond fitting to other statistical matters of
importance in using the models, such as quantification of the
uncertainty of estimation and the formulation of predictive
distributions. These are closely tied to the probability properties of
the RBM model for explaining data generation and our exposition intends
to contribute to a better understanding in this direction.

@@

Let \(\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)\) represent
the states of the visible and hidden nodes in a RBM for some integers
\(\nv, \nh \ge 1\). Each single binary random variable, visible or
hidden, will take its values in a common coding set \(\mathcal{C}\),
where we allow (one of) two possibilities for the coding set,
\(\mathcal{C}=\{0,1\}\) or \(\mathcal{C}=\{-1,1\}\), with ``\(1\)''
always indicating the ``high'' value of the variable. While
\(\mathcal{C}=\{0,1\}\) may be a natural starting point, we argue in
Section \ref{explorations-of-model-properties-through-simulation} that
the coding \(\mathcal{C}=\{-1,1\}\) induces more interpretable model
properties for the RBM. A standard parametric form for probabilities
corresponding to a potential vector of states,
\(X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)\), for the nodes is
\begin{align}
\label{eq:pmf}
f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv} 
\end{align} where
\(\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}\)
denotes the vector of model parameters and the denominator \[
\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
\] is the normalizing function that ensures the probabilities
\eqref{eq:pmf} sum to one. For
\(\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}\)
and \begin{align}
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eq:t}
\end{align} let
\(\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}\)
be the set of possible values for the vector of variables needed to
compute probabilities \eqref{eq:pmf} in the model, and write
\(Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j\)
for the ``neg-potential'' function. The RBM model is parameterized by
\(\boldsymbol \theta\) containing two types of parameters, main effects
and interaction effects. The main effects parameters
(\(\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}\))
weight the values of the visible \(v_i\) and hidden \(h_j\) nodes in
probabilities \eqref{eq:pmf} and the interaction effect parameters
(\(\theta_{ij}\)) weight the values of the connections \(v_i h_j\), or
dependencies, between hidden and visible layers.

@@

RBMs have risen to prominence in recent years due to their connection to
deep learning (see Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}; Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2012efficient}{2012}; Srivastava,
Salakhutdinov, and Hinton
\protect\hyperlink{ref-srivastava2013modeling}{2013} for examples). By
stacking multiple layers of RBMs in a deep architecture, proponents of
the methodology claim to produce the ability to learn ``internal
representations that become increasingly complex, which is considered to
be a promising way of solving object and speech recognition problems''
(Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009}, 450). The stacking
is achieved by treating a hidden layer of one RBM as the visible layer
in a second RBM, and so on, until the desired multi-layer architecture
is created.

@@

In Random Graph Model theory, \emph{model degeneracy} means there is a
disproportionate amount of probability placed on only a few elements of
the sample space, \(\mathcal{X}\), by the model (Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). For random graph
models, \(\mathcal{X}\) denotes all possible graphs that can be
constructed from a set of nodes and an exponentially parameterized
random graph model has a distribution with a probability mass function
of the form \begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*} where
\(\boldsymbol \theta \in \Theta \subset \mathbb{R}^q\) is the model
parameter, and \(\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q\)
is a vector of statistics based on the adjacency matrix of a graph.
Here, as earlier for RBMs,
\(\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)\)
is the normalizing function. Let \(C\) denote the convex hull of the
potential outcomes of sufficient statistics,
\(\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}\),
under the model above. Handcock
(\protect\hyperlink{ref-handcock2003assessing}{2003}) classifies an
exponentially parameterized random graph model at \(\boldsymbol \theta\)
as \emph{near-degenerate} if the mean value of the vector of sufficient
statistics under \(\boldsymbol \theta\),
\(\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\),
is close to the boundary of \(C\). Intuitively, if a model is
near-degenerate in the sense that only a small number of elements of the
sample space \(\mathcal{X}\) have positive probability, the expected
value \(\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\) is
an average of that same small number of values of \(t( \boldsymbol x)\)
(defining the boundary of the hull \(C\)) and can be expected to
\emph{not} be pulled deep into the interior of \(C\).

A RBM model can be thought to exhibit an analogous form of
\emph{near-degeneracy} when there is a disproportionate amount of
probability placed on a small number of elements in the sample space of
visibles and hiddens, \(\mathcal{C}^{\nv + \nh}\). Using the idea of
Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003}), when the
random vector
\(\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}\)
from \eqref{eq:t}, appearing in the neg-potential function
\(Q_{\boldsymbol \theta}(\cdot)\), has a mean vector
\(\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}\)
close to the boundary of the convex hull of \(\mathcal{T}\), and the RBM
model can be said to exhibit near-degeneracy at
\(\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}\). Here the
mean of \(\boldsymbol t(\boldsymbol x)\) is \begin{align*}
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  &= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
\end{align*}
\hypertarget{instability}{%
\subsubsection{Instability}\label{instability}}

@@

S-unstable RBM model sequences are undesirable for several reasons. One
is that, as mentioned above, small changes in data can lead to
overly-sensitive changes in probability under the data model. Consider,
for example, \[
\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(\boldsymbol v)}{P_{\boldsymbol \theta_\nv}(\boldsymbol v^*)} : \boldsymbol v \text{ }\& \text{ } \boldsymbol v^* \in \mathcal{C}^\nv \text{ differ in exactly one component}\right\},
\] denoting the biggest log-probability ratio for a one-component change
in data outcomes (visibles) at a RBM parameter
\(\boldsymbol \theta_\nv\). We then have the following result.

@@

marginal mean-structures fail to hold as anticipated by consideration of
a model statement. We adapt this notion to RBM models. Note that in a

@@

destroying simple interpretation of dependence parameters
\(\theta_{ij}\), \(j=1,\ldots,\nh\) as local conditional modifications
of an overall marginal mean structure \(\theta_{v_i}\), as appearing for
